Supplementary material for -- RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R

### 2014-07-11 manuscripts/manuscript.Rmd
update runs of manuscript

output: 
  pdf_document:
    template: components/elsarticle.latex
    keep_tex: false
    fig_caption: true

@@

  - Metadata
    - Adding basic metadata
    - Adding taxonomic identifers
    - Adding Generic metadata
    - Parsing metadata. XML. RDF.
    - Extending NeXML with metadata: Example defining simmap stochastic character mappings.

@@

<!-- Need to say some of this stuff without repeating too much of existing literature -->

@@

<!-- FIXME -->

@@

The ever expanding needs of the phylogenetics community are rapidly
out-pacing the capacity of current standards, which is resulting in a
radiation of different data representations and a loss of interoperability
at the very time when the growth of available data and mehtods has made
that interoprability most valuable.

For example, the currently popular NEXUS file format cannot consistently
represent horizontal gene transfer or ambiguity in reading a character
(such as a DNA sequence base pair).  This has led developers to
repeatedly add small modifications to the standard to meet the needs
of their own software, which results in mutually incompatible dialects
that largely defeat the advantages of a shared  file format.
NeXML was developed to addresses each of these problems `r citep("Vos NeXML Systematic Biology")`  
by drawing on robust web standards in data representation.

@@

The R lanugage `r citep(citation())` has become an important enviroment for 
phylogenetics research, with over 30 packages available including several
with over 1000 citations `r citep("http://cran.r-project.org/web/views/Phylogenetics.html")`,
though stand-alone programs remain an important part of this ecosystem
for tasks such as generating phylogenetic trees from sequence data.  Within 
this ecosystem a variety of different data formats for
phylogenetic data have been developed and modified. Most but not all 
are based on the simple and popular `phylo` format provided by the
ape package `r citep(citation("ape"))`.

@@

In this paper, we introduce the reader to the concepts and use of
RNeXML and related technologies. We begin by discussing methods to read
phylogenetic tree data and character trait data in the NeXML format into
R, and also how to write phylogenetic data and character trait data into
NeXML. The heart of this section focuses on tools for reading and writing
the kind of semantically rich metadata that most sets NeXML apart from
other phylogenetic data formats. Starting simple, we use examples of
increasing complexity to illustrate concepts and tools from biodiversity
informatics that allow users to express and interpret complicated data
structures automatically.  We end with an example that uses metadata to
extend the NeXML standard to describe stochastic character mapping, one
of the kinds of data annotation that is currently available in R formats
but lacks a consistent, extensible export format. We illustrate how NeXML
files can be formally published to a data respository directly from R.
Finally, we provide an brief description of how the software has been
implemented, what quality control measures are in place, as well as
information about software reuse and getting support.

@@

A NeXML document can be thought of as having multiple components: (1) phylogeny
data (2) character-trait data, (3) operational taxonomic units, and (4) metadata. 
`RNeXML` seeks to provide a seemless way to convert these to and from their 
corresponding R object representations, such as `ape::phylo` phylogenies, character
matrices (`data.frames`) and metadata lists (or for advanced users,
`XMLInternalDocument` and `rrdf` triplestores). Metadata can provide
information about the study in which the phylogeny was generated or
applied (such as authors, citations to publications, timestamps, and even
descriptions of methods), but can also be used to annotate individual
data components.  We illustrate how metadata annotations can be added to
the operational taxonomic units of the phylogeny and can be used to extend
the data representation to new uses such as stochastic character maps.

@@

<!-- have we already said all this above?-->  <!-- no, seems appropriate to repeat here, -scott --> The ability to read in
phylogenetic trees in the NeXML format opens up a wide and rapidly
growing array of resources to the R user.  For instance, a user can access
all the phylogenies available in TreeBASE through the `nexml` format.
`RNeXML` can read directly from a URL:

@@

Previously, this was possible using the `treebase` package `r citep("Boettiger Temple Lang treebase 2012")`.
for R, which relied on the NEXUS parser provided in the `ape` package.
Due to inconsistencies present in the NEXUS standard `r citep("Vos NeXML 2012")`, 
a significant number <!-- State explicitly --> of TreeBASE NEXUS
files could not be read in this way.  Using the more precisely defined
NeXML standard, every valid NeXML document on TreeBASE can now successfully
be parsed to R objects.

@@

Returns a `data.frame` with columns as characters and rows as
taxa. Continuous and discrete characters are represented as seperate
blocks in NeXML files, but will be combined as seperate columns of a
single data frame by `get_characters()` if they correspond to the same
taxa. The `get_characters_list()` function will consistently receive
a list of data frames corresponding to each character matrix block of
the NeXML.

@@

<!-- in the Perl library I also embed the value of the environment variable
$USER as a document-level annotation -->
`RNeXML` automatically adds basic metadata by default such as a timestamp
of when the tree was produced and an open data license (CC0, compatible
with deposition on Dryad data repository), which can be configured
as needed.

@@

`RNeXML` provides various functions to assist in automatically adding
other more specific metadata to various elements of the NeXML data.
One common use case is that of taxonomic identifers.  Misspellings or
inconsistent use of taxonomic names in phylogenetic data are a common
source of frustration in reusing phylogenetic trees, especially when 
integrating data from different sources: trees might come from one place
that uses one taxonomy, character data and occurrence data might come from
other places that use different taxonomies. Hence, not just misspellings, 
but also synonyms, homonyms and incompatible taxon concepts might plague
the end user. By associating the
taxonomic labels against a specific, centralized database we can catch
some of these inconsistencies and indicate more precisely which classification
we are appealing to in the event of conflicting taxonomic definitions.
In this example, we indicate that we intend the bird orders listed in
the tree to correspond to the NCBI definitions

@@

So far our examples have relied on functions that have a built-in
understanding of the kind of metadata being provided, and can thus
generate the appropriate metadata with little further guidence.
In order to be machine-readable and understandable, NeXML requires
all metadata annotations follow the RDFa format (see @Vos_2012) which
explicitly declares the meaning of the annotation in a way a computer
can understand.  For instance, the title we gave to the `bird.orders`
phylogeny appears in the NeXML as:

@@

This URL links us to a webpage providing both a human and machine-readable
version of precisely what the term `title` means, showing us that it
refers to `title` in the sense of a creative work, rather than the title
of a person, a legal title, the verb title, etc.  This precision can allow
us to perform precise computer reasoning on metadata, without relying on
heuristics or assumptions, as we illustrate later.  For the moment though,
the need to refer to an explicit controlled vocabulary (or ontology)
with a url, saying "dc:title" rather than just saying the property is
"title", is something of a burden in creating metadata.  The previous
examples involve functions that already know what kind of metadata is
being provided (a title, a data creator, etc), and so can provide the
appropriate vocabulary automatically.  The main advantage of linking to
an external vocabulary, however, is the ability to express concepts that
are not already "built in" to the data model.  `RNeXML` provides tools
to generate arbitrary metadata annotations given an external vocabulary,
such as we illustrate here.

@@

We can add this annotation to our existing `birds` nexml file using the
`add_meta` function.  Because we do not specify a level, it is added to
the root node, refering to the nexml file as a whole.

@@

Because `skos` is not in the current namespace list, we add it with a
url when adding this meta element.  We also specify that this annotation
be placed at the level of the `trees` subnode in the NeXML file.

@@

<!-- does this mean we can use CURIEs whose prefix has not yet been 
bound to a namespace? that would be bad. there should probably be an
error message when that happens -->
If we forget to add the namespace with add_meta, we can always add it
later using the `add_namespaces` function,

```{r}
birds <- add_namespaces(c(skos = "http://www.w3.org/2004/02/skos/core#"), birds)
```

Note that the `add_namespaces` function will automatically avoid creating
any duplicate definitions. 
For finer control of the level at which a `meta` element is added,
we will manipulate the nexml R object directly using S4 subsetting,
as shown later.

<!--FIXME Ugh. y'all have much more practice explaining this well and without tangents, 
so please feel free to edit or re-write.  -->
<!-- I am not quite sure what points we are trying to make here -->

@@

A call to the nexml object prints some metadata summarizing the data structure:

@@

and can ask for a particular element using the standard list subsetting
mechanism (i.e. either the name of an element or its numeric position),

@@

<!-- I feel like we already covered what dc:title is -->
Following the link we can see that `dc` refers to the "Dublin Core"
vocabulary, and the term `title` refers to the title of a creative work
(as opposed to the title of a person, a legal title, the verb title, etc).
Such explicit definitions allow a computer to reason about these objects,
as we shall see later.

@@

This approach does not preserve the relationship between the metadata element and the 
`otu` it was annotating, and as such is most useful for quick visual exploration. 
Metadata values for specific elements or sets of elemetns can always be accessed using
the S4 subsetting mechanisms (see section below).

@@

Consider a better example than this one:

```{r}
rdf <- get_rdf(birds)
```

The most powerful way to take advantage of the machine-readable nature
of the metadata involves the construction of SPARQL queries. This
approach can exploit the reasoning and logical deduction capacity of
formal vocabularies (ontologies). For a proper introduction to SPARQL
queries, see `r citet("Hartig Introduction to SPARQL")`.  Meanwhile, we just introduce the syntax
necessary to apply such queries to NeXML files from R.  First we load
the necessary R library `r citep(citation("rrdf"))` and import the RDF-extracted metadata:

```{r}
library("rrdf")
saveXML(rdf, "rdf_meta.xml") # rrdf requires a file name, so we must write the XML out first
lib <- load.rdf("rdf_meta.xml")
```

<!-- FIXME it would be nice if we had a more persuasive example here. directly getting the value
of some predicate can be done the syntax-oriented way by treating the data as XML. what I
think we should be going for here is an example of a slightly more complex query that would
be painful to perform by scanning the XML document. I'm thinking of some sort of join of
multiple predicate values. Nothing great comes to mind yet, though... ->

We can then make arbitrary SPARQL queries directly from R.  For instance,
here is a simple query to extract the value of any `dc:title` element, analogous
to the XPath expression above:

```{r}
sparql.rdf(lib, "SELECT ?title WHERE { ?x <http://purl.org/dc/elements/1.1/title> ?title}")
```

@@

Thus the the annotation for an edge that switches from state `s2` to state 
`s1` of character `cr1` would be constructed like this:

@@

<!-- How could this be improved? Should we use something more formal? 
We want this example to be something other users could reasonably
do themselves without expertise in RDFa or XML, but also to be a good
model case that doesn't cut corners. @cboettig -->

<!-- I don't see how this could be improved much without forcing people
to fire up protege (let's not!). As it happens, the Dublin Core URI, 
for example, points to an HTML page so even in that case automatic 
reasoners would not be able to consume the vocabulary and draw inferences
about, e.g. identity of predicates (owl:sameAs and such). @rvosa -->

Using this convention we can generate NeXML files containing simmap
data, read those files into R, and convert them back into the `phytools`
package format. These simple functions serve as further illustration of
how `RNeXML` can be used to extend the NeXML standard.  We illustrate
their use briefly here, starting with  loading a nexml object containing
a simmap reconstruction into R:

@@

Likewise, we can convert the object back in the nexml format and write
it out to file to be read by other users.

@@

_Show downloading the nexml file directly from figshare, explain value
of global paths instead of personal directory paths appearing in scripts, etc_

@@

<!-- Wouldn't it be nice to publish to Dryad or TreeBASE in this way?  -->

@@

[^3]: By using the commenting mechanism of the Newick format, it is
possible that other software that doesn't also use the comment mechanism
for some other such purpose would be able to successfully parse the tree.
However there is no way to guarentee that this is the case or for the
data format to describe its use.

>>>>

output: 
  pdf_document:
    template: components/elsarticle.latex
    keep_tex: true
    fig_caption: true

@@

  - Metadata
    - Adding basic metadata
    - Adding taxonomic identifiers
    - Adding Generic metadata
    - Parsing metadata. XML. RDF.
    - Extending NeXML with metadata: Example defining simmap stochastic character mappings.

@@

@@

<!-- FIXME

@@

-->

The ever expanding needs of the phylogenetics community are rapidly
out-pacing the capacity of current standards, which is resulting in a
radiation of different data representations and a loss of interoperability
at the very time when the growth of available data and methods has made
that interoperability most valuable.

For example, the currently popular NEXUS file format cannot consistently
represent horizontal gene transfer or ambiguity in reading a character
(such as a DNA sequence base pair).  This has led developers to
repeatedly add small modifications to the standard to meet the needs
of their own software, which results in mutually incompatible dialects
that largely defeat the advantages of a shared  file format.
NeXML was developed to addresses each of these problems `r citep("Vos NeXML Systematic Biology")`  
by drawing on robust web standards in data representation.
The design and expressiveness of NeXML makes it much easier to meet
the rising bar for best practices in data sharing in phylogenetic 
research `r citep("10.1371/currents.tol.bf01eff4a6b60ca4825c69293dc59645")`.

@@

The R language `r citep(citation())` has become an important environment for 
phylogenetics research, with over 30 packages available including several
with over 1000 citations `r citep("http://cran.r-project.org/web/views/Phylogenetics.html")`,
though stand-alone programs remain an important part of this ecosystem
for tasks such as generating phylogenetic trees from sequence data.  Within 
this ecosystem a variety of different data formats for
phylogenetic data have been developed and modified. Most but not all 
are based on the simple and popular `phylo` format provided by the
ape package `r citep(citation("ape"))`.

@@

In this paper, we introduce the reader to the concepts and use of
RNeXML and related technologies. We begin by discussing methods to read
phylogenetic tree data and character trait data in the NeXML format into
R, and also how to write phylogenetic data and character trait data into
NeXML. The heart of this section focuses on tools for reading and writing
the kind of semantically rich metadata that most sets NeXML apart from
other phylogenetic data formats. Starting simple, we use examples of
increasing complexity to illustrate concepts and tools from biodiversity
informatics that allow users to express and interpret complicated data
structures automatically.  We end with an example that uses metadata to
extend the NeXML standard to describe stochastic character mapping, one
of the kinds of data annotation that is currently available in R formats
but lacks a consistent, extensible export format. We illustrate how NeXML
files can be formally published to a data repository directly from R.
Finally, we provide an brief description of how the software has been
implemented, what quality control measures are in place, as well as
information about software reuse and getting support.

@@

A NeXML document can be thought of as having multiple components: (1) phylogeny
data (2) character-trait data, (3) operational taxonomic units, and (4) metadata. 
`RNeXML` seeks to provide a seamless way to convert these to and from their 
corresponding R object representations, such as `ape::phylo` phylogenies, character
matrices (`data.frames`) and metadata lists (or for advanced users,
`XMLInternalDocument` and `rrdf` triplestores). Metadata can provide
information about the study in which the phylogeny was generated or
applied (such as authors, citations to publications, timestamps, and even
descriptions of methods), but can also be used to annotate individual
data components.  We illustrate how metadata annotations can be added to
the operational taxonomic units of the phylogeny and can be used to extend
the data representation to new uses such as stochastic character maps.

@@

The ability to read in phylogenetic trees in the NeXML format opens up a
wide and rapidly growing array of resources to the R user.  For instance,
a user can access all the phylogenies available in TreeBASE through the
`nexml` format.  `RNeXML` can read directly from a URL:

@@

Previously, this was possible using the `treebase` package `r citep("Boettiger Temple Lang treebase 2012")`
for R, which relied on the NEXUS parser provided in the `ape` package.
Due to inconsistencies present in the NEXUS standard `r citep("Vos NeXML 2012")`, 
a significant number <!-- State explicitly --> of TreeBASE NEXUS
files could not be read in this way.  Using the more precisely defined
NeXML standard, every valid NeXML document on TreeBASE can now successfully
be parsed to R objects.

@@

Returns a `data.frame` with columns as characters and rows as
taxa. Continuous and discrete characters are represented as separate
blocks in NeXML files, but will be combined as separate columns of a
single data frame by `get_characters()` if they correspond to the same
taxa. The `get_characters_list()` function will consistently receive
a list of data frames corresponding to each character matrix block of
the NeXML.

@@

<!-- in the Perl library I also embed the value of the environment variable
$USER as a document-level annotation -->
`RNeXML` automatically adds basic metadata by default such as a time-stamp
of when the tree was produced and an open data license (CC0, compatible
with deposition on Dryad data repository), which can be configured
as needed.

@@

`RNeXML` provides various functions to assist in automatically adding
other more specific metadata to various elements of the NeXML data.
One common use case is that of taxonomic identifiers.  Misspellings or
inconsistent use of taxonomic names in phylogenetic data are a common
source of frustration in reusing phylogenetic trees, especially when
integrating data from different sources: trees might come from one place
that uses one taxonomy, character data and occurrence data might come from
other places that use different taxonomies. Hence, not just misspellings,
but also synonyms, homonyms and incompatible taxon concepts might plague
the end user. By associating the taxonomic labels against a specific,
centralized database we can catch some of these inconsistencies and
indicate more precisely which classification we are appealing to in the
event of conflicting taxonomic definitions.  In this example, we indicate
that we intend the bird orders listed in the tree to correspond to the
NCBI definitions

@@

So far our examples have relied on functions that have a built-in
understanding of the kind of metadata being provided, and can thus
generate the appropriate metadata with little further guidance.
In order to be machine-readable and understandable, NeXML requires
all metadata annotations follow the RDFa format (see @Vos_2012) which
explicitly declares the meaning of the annotation in a way a computer
can understand.  For instance, the title we gave to the `bird.orders`
phylogeny appears in the NeXML as:

@@

This URL links us to a web page providing both a human and machine-readable
version of precisely what the term `title` means, showing us that it
refers to `title` in the sense of a creative work, rather than the title
of a person, a legal title, the verb title, etc.  This precision can allow
us to perform precise computer reasoning on metadata, without relying on
heuristics or assumptions, as we illustrate later.  For the moment though,
the need to refer to an explicit controlled vocabulary (or ontology)
with a url, saying "dc:title" rather than just saying the property is
"title", is something of a burden in creating metadata.  The previous
examples involve functions that already know what kind of metadata is
being provided (a title, a data creator, etc), and so can provide the
appropriate vocabulary automatically.  The main advantage of linking to
an external vocabulary, however, is the ability to express concepts that
are not already "built in" to the data model.  `RNeXML` provides tools
to generate arbitrary metadata annotations given an external vocabulary,
such as we illustrate here.

@@

We can add this annotation to our existing `birds` NeXML file using the
`add_meta` function.  Because we do not specify a level, it is added to
the root node, referring to the NeXML file as a whole.

@@

Because `skos` is not in the current namespace list, we add it with a
url when adding this meta element.  We also specify that this annotation
be placed at the level of the `trees` sub-node in the NeXML file.

@@

For finer control of the level at which a `meta` element is added,
we will manipulate the `nexml` R object directly using S4 sub-setting,
as shown in the supplement.

@@

A call to the `nexml` object prints some metadata summarizing the data structure:

@@

and can ask for a particular element using the standard list sub-setting
mechanism (i.e. either the name of an element or its numeric position),

@@

@@

This approach does not preserve the relationship between the metadata element and the 
`otu` it was annotating, and as such is most useful for quick visual exploration. 
Metadata values for specific elements or sets of elements can always be accessed using
the S4 sub-setting mechanisms (see section below).

@@

The most powerful way to take advantage of the machine-readable nature
of the metadata involves the construction of SPARQL queries. This
approach can exploit the reasoning and logical deduction capacity of
formal vocabularies (ontologies). For a proper introduction to SPARQL
queries, see `r citet("http://www.w3.org/TR/rdf-sparql-query/")` and
`r citet("Hartig Introduction to SPARQL")`.

The real expressiveness of SPARQL is better illustrated through more
complex examples such as the one shown in the supplementary materials
(which can also be run from R using the command: `demo("sparql", "RNeXML")`).
This demonstration constructs a cladogram based on the taxonomic ranks
provided in the NeXML. SPARQL allows us to make explicit use of ontological
concepts such as manipulating sub-classes (i.e. that "Genus" is a subclass 
of "Family"), or joining multiple predicates (i.e. retrieve the NCBI identifier
of the OTU that has taxon rank of "Order").

@@

Thus the annotation for an edge that switches from state `s2` to state 
`s1` of character `cr1` would be constructed like this:

@@

Using this convention we can generate NeXML files containing simmap
data, read those files into R, and convert them back into the `phytools`
package format. These simple functions serve as further illustration of
how `RNeXML` can be used to extend the NeXML standard.  We illustrate
their use briefly here, starting with  loading a `nexml` object containing
a simmap reconstruction into R:

@@

Likewise, we can convert the object back in the NeXML format and write
it out to file to be read by other users.

@@

@@

@@

[^3]: By using the commenting mechanism of the Newick format, it is
possible that other software that doesn't also use the comment mechanism
for some other such purpose would be able to successfully parse the tree.
However there is no way to guarantee that this is the case or for the
data format to describe its use.

### 2014-07-11 manuscripts/supplement.Rmd
update runs of manuscript

>>>>

```{r supplement-compile-settings, include=FALSE}
library("methods")
library("knitr")
opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, 
               cache = 1, comment = NA, verbose = TRUE)
basename <- 'supplement' 
opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
               cache.path = paste("components/cache/", basename, "/", sep=""))

library("knitcitations")
cleanbib()
cite_options(citation_format = "pandoc", check.entries=FALSE)
```

## SPARQL Queries

```{r}
library(rrdf)
library(phytools)
library(RNeXML)
```

```{r}
nexml <- nexml_read(system.file("examples/primates.xml", package="RNeXML"))
```

Extract the RDF graph from the nexml

```{r}
rdf <- get_rdf(system.file("examples/primates.xml", package="RNeXML"))
saveXML(rdf, "rdf_meta.xml") # rrdf requires a file name, so we must write the XML out first
graph <- load.rdf("rdf_meta.xml")
```

fetch the NCBI URI for the taxon that has rank 'Order', i.e. the root of the primates. The dot operator `.` between clauses implies a join, in this case

```{r}
root <- sparql.rdf(graph, 
    "SELECT ?uri WHERE { 
        ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#rank> <http://rs.tdwg.org/ontology/voc/TaxonRank#Order> . 
        ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> ?uri    
    }"               
)
```

Define a function to get the name

```{r}
get_name <- function(id) {
    max <- length(nexml@otus[[1]]@otu)
    for(i in 1:max) {
        if ( nexml@otus[[1]]@otu[[i]]@id == id ) {
            label <- nexml@otus[[1]]@otu[[i]]@label
            label <- gsub(" ","_",label)
            return(label)
        }
    }
}
```

 define a recursive function to build newick

```{r}
recurse <- function(node){

    # fetch the taxonomic rank and id string
    rank_query <- paste0(
        "SELECT ?rank ?id WHERE {
            ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> <",node,"> .
            ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#rank> ?rank
          }")
    result <- sparql.rdf(graph, rank_query)

    # get the local ID, strip URI part
    id <- result[2]
    id <- gsub("^.+#", "", id, perl = TRUE)

    # if rank is terminal, return the name
    if (result[1] == "http://rs.tdwg.org/ontology/voc/TaxonRank#Species") {
        return(get_name(id))
    }

    # recurse deeper
    else {
        child_query <- paste0(
            "SELECT ?uri WHERE {
                ?id <http://www.w3.org/2000/01/rdf-schema#subClassOf> <",node,"> .
                ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> ?uri
            }")
        children <- sparql.rdf(graph, child_query)

        # the newick can be made to contain interior node labels by inserting get_name(id), before the sep="" argument
        return(paste("(", 
                     paste(sapply(children, recurse), 
                           sep = ",", collapse = "," ), 
                     ")",  sep = "", collapse = ""))
    }
}
```

build the tree and visualize it

```{r}
newick <- paste(recurse(root), ";", sep = "", collapse = "")
tree <- read.newick(text = newick)
collapsed <- collapse.singles(tree)
plot(collapsed, type = "cladogram")
```
